Modelling fine-scale dive behavior with hidden Markov models

### 2020-07-07 Report/Sections/appendix.tex
Updates from Overleaf

From here it is straightforward to calculate both $Z^{*(1)}$ and $Z^{*(2)}$ after setting $\tilde{f}$. We pick $\tilde{f} = 10$ periods per window, or 5 hertz. Note that $Z^{*(1)}_{t,s^*} = \mathcal{R}\left(\hat{Y}^{(0)}_{t,s^*}\right)$, so selecting $\sigma^{*(i)}$ and $\phi^{*(1)}$ appropriately will ensure the distributions from (eqn \ref{}).

@@

$$Z^{*(2)} = \sum_{k=1}^{10} b^{(k)}_{t,s^*}$$
$$Z^{*(2)} \sim {\rm{Gamma}}

>>>>

From here it is straightforward to calculate both $Z^{*(1)}$ and $Z^{*(2)}$ after setting $\tilde{f}$. We pick $\tilde{f} = 10$ periods per window, or 5 hertz. Note that $Z^{*(1)}_{t,s^*} = \mathcal{R}\left(\hat{Y}^{(0)}_{t,s^*}\right)$, so selecting $\sigma^{*(i)}$ and $\phi^{*(1)}$ appropriately will ensure correct distribution.

@@

$$Z^{*(2)} = \sum_{k=1}^{10} b^{(k)}_{t,s^*}$$
$$Z^{*(2)} \sim {\rm{Gamma}}$$

### 2020-07-07 Report/Sections/intro.tex
Nancy Revisions
Updates from Overleaf

@@

The field of animal movement is in the midst of a ``data renaissance" in which advancements in tagging technology have given rise to an explosion of data available for statistical modeling. In response, researchers have introduced a variety of new statistical techniques to infer animal behavior from movement data \cite{Hooten:2017}.

These techniques usually fall into one of two different categories: continuous-time models and discrete-time models. Continuous-time models often model the dynamics of an animal as the solution to some stochastic differential equation and can incorporate observations taken at irregular time intervals. These methods are often computationally intractable and require approximate inference techniques such as Markov-chain Monte Carlo (MCMC) methods to perform inference. On the other hand, discrete-time techniques use summary statistics taken at discrete time steps to model the animal's behavior and require that observations be taken at regular time intervals. The primary advantage of discrete-time methods is that the associated likelihood is relatively easy to compute, facilitating fast maximum likelihood estimates for parameters.

One of the most prevalent discrete-time techniques in the literature of late is the hidden Markov model (HMM), where observations depend upon the state of an associated unobserved behavioral process following Markovian dynamics \cite{Patterson:2017}. Importantly, under the traditional HMM model, subsequent observations are assumed to be independent from one another conditioned on the underlying behavioral process. However, this assumption is often violated in the real world processes. For example, the location of an animal at a given time is highly correlated with the location of that animal one second later. Several publications have dealt with this issue in the past, including the hidden movement Markov model (HMMM) by Whoriskey et al \cite{Whoriskey:2016}.

\textit{The conditionally autoregressive hidden Markov model (CarHMM): Inferring Behavioral States from animal tracking data exhibiting conditional autocorrelation} by Ethan Lawler, Kim Whoriskey, William H. Aeberhard, Chris Fields, and Joanna Mills Flemming introduces a new framework to model autocorrelation in hidden Markov models. The paper also recommends several best practices for data exploration, model selection, and data preprocessing. This paper summarizes the findings of Lawler et al, elaborates on some of the findings and recommendations within, and draws a connection between the CarHMM and a continuous-time method presented by Michelot et al \cite{Michelot:2019}. The CarHMM is also adapted and applied to real-world dive data to analyze the intra-dive behavior of a killer whale off the coast of British Columbia, Canada.

>>>>

%\section{Introduction}

The field of animal movement is in the midst of a ``data renaissance" where advancements in tagging technology have given rise to an explosion of data available for statistical modeling. In particular, tagging technologies are capable of recording observations at rates of tens of hertz, resulting in time series containing millions of observations over the course of several hours. In response, researchers have introduced a variety of new statistical techniques to infer animal behavior from movement data \citep{Hooten:2017}.

One of the most prevalent techniques in the literature of late is the hidden Markov model (HMM), where observations depend upon the state of an associated unobserved behavioral process following Markovian dynamics \citep{Patterson:2017}. Importantly, under the traditional HMM model, subsequent observations are assumed to be independent from one another conditioned on the underlying behavioral process. However, this assumption is often violated in the real world processes, especially when observations are taken a high frequencies. For example, the location of an animal at a given time is highly correlated with the location of that animal one second later. Several publications have dealt with this issue in the past, including the hidden movement Markov model (HMMM) by Whoriskey et al \citep{Whoriskey:2016} and the conditionally auto-regressive hidden Markov model (CarHMM) by Lawer et al. \citep{Lawler:2019}. The CarHMM in particular explicitly models auto-correlation into an HMM while maintaining the structure needed to run the forward algorithm. It also only adds one additional parameter per possible hidden state

Another issue that arises in high-frequency data is that several simultaneous behavioral processes may occur at different time scales. In this work we consider the example where killer whales perform a variety of different types of dives at a coarse scale, but also exhibit many different types of swimming behaviors within dives at a fine scale. One solution to this issue is to use a hierarchical hidden Markov model (HHMM) \citep{Barajas:2017} \citep{Adam:2019}. HHMMs model the entire time series in question as a nested structure of hidden Markov models (HMM) where each HMM corresponds to one behavioral process.

At the shortest time scales, however, observations often exhibit complicated dependence structures which cannot be easily captured by traditional HMMs, CarHMMs, or HHMMs. Examples included periodic fluking behavior in killer whales off the coast of Vancouver, BC, and swimming patterns of horn sharks of the coast of Southern California \citep{Adam:2019}. One solution is to model the fine-scale behavior using a continuous-time model, which usually involves modelling the dynamics of an animal as the solution to a stochastic differential equation. Continuous-time models are more flexible than their discrete-time counterparts and can incorporate observations taken at irregular time intervals. However, they are often computationally intractable and require approximate inference techniques such as Markov-chain Monte Carlo (MCMC) methods to perform inference.

This work investigates how to incorporate fine-scale processes into the larger structure of hierarchical hidden Markov models while maintaining computational efficiency. We describe a general procedure that can be used to extract features from highly structured fine-scale behaviors that otherwise could not be modeled with existing HMM models. In addition, we bridge the gap between the discrete CarHMM and certain continuous-time stochastic process models by showing that the two are equivalent under certain conditions. We then perform a simulation study to compare the performance each existing model with ours in a controlled setting. Finally, we apply our method to dive data collected from a Northern resident killer whale off the coast of British Columbia, Canada.

We consider several models- the classical hierarchical hidden Markov model (HHMM) \citep{Barajas:2017}, the CarHMM model of \citep{Lawler:2019}, and our new model that blends the two (CarHHMM). The HHMM consists of a coarse-scale process and a fine-scale process, with standard HMMs modelling both the coarse- and fine-scale processes. The CarHMM  modifies the usual HMM by modelling additional dependence into the sequence of observed data. In our combination of the HHMM and the CarHHMM models, we keep the usual Markov structure for the coarse-scale process, but use the CarHMM model for the fine-scale process.

%%%%%%%%%% Nancy's comments %%%%%%%%%%%%%

\iffalse

@@

{\bf{NH: These are my comments.  I hope they help you.}}

This is where you describe how your method contributes to statistical methodology. That is the big goal.  So there will be literature review of directly relevant work.  You don't want to leave the non-expert completely lost though, so it is a balance, how much background you give.
Thank goodness for the Zucchini book, a great place to direct readers for further background.  
The introduction is a good place to put information about the motivating data set and how researchers have modelled it, what the shortcomings have been.  I see that you have some great statements about this in your Background section.  Those statements can be moved to the introduction (with editing -- you can see how I edited).

A hard part about the introduction is that you want to use terminology like CarHMM without getting into too many technical details.  Sometimes you just have to be technical.  But I think you can avoid a lot of that.

Here is a great thing you wrote that would go well in the introduction, with some editing so that it doesn't rely on notation that hasn't been introduced. You could introduce the basic HMM notation in the intro, if you need to.  I don't think introducing HMM notation is too technical.
\begin{quote}
 One of the key assumptions of both HMMs and HHMMs is \textit{conditional independence} between observations on both the coarse and fine scale. Namely, when given the state $X_t$, $Y_t$ is assumed to be independent from all other observations, including $Y_{t-1}$. Therefore, traditional HMMs and HHMMs can fail when the observations $Y$ exhibit significant correlation in time.
 [leave in these examples for intro]  Examples include fluking in marine mammals in Vancouver, BC (see the results section) and the swimming behavior of horn sharks off the coast of Southern California \citep{Adam:2019}.

One way to deal with auto-correlation in fine-scale behavioral processes is to use the CarHMM, or \textit{conditionally auto-regressive hidden Markov model}, introduced by Lawler et al \citep{Lawler:2019}. .....  The CarHMM therefore explicitly models auto-correlation into the emission distributions of the HMM while maintaining the structure needed to run the forward algorithm. It also only adds one additional parameter per possible hidden state
$(\phi^{(i)})$. 
\end{quote}

\fi

### 2020-07-07 Report/Sections/model.tex
Updates from Overleaf

\section{Model Formulation}

\subsection{Short-time Fourier Transform on Fine-scale processes}

One issue with the CarHMM is that it assumes Markovian dynamics when conditioned on the hidden state, i.e. that any observation $Y^*_{t,t^*}$ depends only on the behavioral state $X^*_{t,t^*}$ and $Y^*_{t,t^*-1}$. However, there are many animal movement processes which violate this Markov property on very fine scales. For example, swimming behaviour of marine mammals is often exhibits periodic since the animal repeatedly flukes (swims up and down) to propel itself forward. Work has been done in the past to model non-Markovian dynamics in the \textit{behavioural} process $X^*_t$ \cite{Langrock:2012}, but to our knowledge addressing non-Markovian dynamics within the observation process $Y^*_t$ is still a relatively unstudied area \textcolor{red}{(MAYBE?)}. With improvements in tagging technology allowing for data to be collected at very high frequencies, noisy and non-Markovian fine scale behavior is likely to persist.

@@

\begin{figure}[h!]
	\centering
	\includegraphics[width=5.5in]{../Plots/fourier_transform.png}
	\caption{Visualization of transforming $Y^*_t$ into $Z^*_t$ using a sliding window and Fourier transform.}
	\label{fig:fourier_example}
\end{figure}

Note that it is possible to accommodate for unequal time steps within each window by using the \textbf{non-uniform discrete Fourier transform (NDFT)}. We do not describe the details of this method in this work, but the generalization is straightforward. Refer to Bagchi et al \cite{Bagchi:2001} for details.

Fourier analysis has previously been used in the field of animal movement to explain animal behavior \cite{Fehlmann:2017} and specifically fluking \cite{Shorter:2017} from accelerometer data, and Fourier analysis within an HMM has been used to describe daily behavioral cycles of marine mammals  \cite{Heerah:2017}. However, we are currently unaware of any studies which have incorporated Fourier analysis of accelerometer data within the structure of an HMM.

\subsection{Model Structure: combining the HHMM and CarHMM}

Hierarchical hidden Markov models can be used to jointly model simultaneous coarse-scale and fine-scale processes. However, as mentioned before, the fine-scale process $Y^*$ can often exhibit auto-correlation and intricate structure. Transforming $Y^*_t$ to $Z^*_t$ removes fine-scale periodic behavior, but $Z^*_t$ can still exhibit auto-correlation, especially in $Z_{t,s^*}^{*(1)}$. Therefore, we replace the fine-scale HMM within the hierarchical HMM with a CarHMM as shown in (fig \ref{fig:CarHHMM}).
%
\begin{figure}[h!]
	\centering
	\includegraphics[width=5in]{../Plots/CarHHMM.png}
	\caption{Graphical representation of a CarHHMM. The additional arrows representing auto-correlation between observations are shown in red for emphasis.}
	\label{fig:CarHHMM}
\end{figure}
%
Naturally, we refer to this model as the \textbf{CarHHMM}, and it can be implemented both with and without the short-time Fourier-transform (STFT). The likelihood of this model is still easy to calculate using the forward algorithm:
%
$$\calL_{\text{CarHHMM}}(y,z^*;\Theta,\Theta^*,\Gamma,\Gamma^*) = \delta P(y_1,z_1^*;\Theta,\Theta^*,\Gamma^*) \prod_{t=2}^T \Gamma P(y_t,z_t^*;\Theta,\Theta^*,\Gamma^*) \mathbf{1}_N$$
%
where:
%
\begin{align*}
P(y_t,z_t^*;\Theta,\Theta^*,\Gamma^*)  = \text{diag}\Big[&f^{(1)}(y_t)\calL_{\text{CarHMM}}\left(z_t^*;\Theta^{*(1)},\Gamma^{*(1)}\right), \ldots , \\
&f^{(N)}(y_t)\calL_{\text{CarHMM}}\left(z_t^*;\Theta^{*(N)},\Gamma^{*(N)}\right) \Big]
\end{align*}
%

>>>>

%\section{Model Formulation}

\subsection{Moving windows and the Short-time Fourier transform}
\label{subsec:STFT}

One issue with the CarHMM is that it assumes Markovian dynamics when conditioned on the hidden state, i.e. that any observation $Y^*_{t,t^*}$ depends only on the behavioral state $X^*_{t,t^*}$ and $Y^*_{t,t^*-1}$. However, there are many animal movement processes which violate this Markov property on very fine scales. For example, swimming behaviour of marine mammals is often exhibits periodic since the animal repeatedly flukes (swims up and down) to propel itself forward. Work has been done in the past to model non-Markovian dynamics in the \textit{behavioural} process $X^*_t$ \citep{Langrock:2012}, but to our knowledge addressing non-Markovian dynamics within the observation process $Y^*_t$ is still a relatively unstudied area \textcolor{red}{(MAYBE?)}. With improvements in tagging technology allowing for data to be collected at very high frequencies, noisy and non-Markovian fine scale behavior is likely to persist.

@@

\begin{figure}[ht]
	\centering
	\includegraphics[width=5in]{../Plots/fourier_transform.png}
	\caption{Visualization of transforming $Y^*_t$ into $Z^*_t$ using a sliding window and Fourier transform.}
	\label{fig:fourier_example}
\end{figure}

It is possible to accommodate for unequal time steps within each window by using the \textbf{non-uniform discrete Fourier transform (NDFT)}. We do not describe the details of this method in this work, but the generalization is straightforward. Refer to Bagchi et al \citep{Bagchi:1999} for details.

The advantages of using Fourier analysis within an HMM has been recently demonstrated in the context of describing daily behavioral cycles of marine mammals \citep{Heerah:2017}. In addition, Fourier analysis has previously been used in the field of animal movement to explain animal behavior \citep{Fehlmann:2017} and specifically fluking \citep{Shorter:2017} from accelerometer data. Thus, incorporating Fourier analysis of accelerometer data within the structure of an HMM appeared a promising simple approach to account for additional correlation in data that is cyclical in nature.

\subsection{Model Structure: combining the HHMM and CarHMM}
\label{subsec:model_structure}

Hierarchical hidden Markov models can be used to jointly model simultaneous coarse-scale and fine-scale processes. However, as mentioned before, the fine-scale process $Y^*$ can often exhibit auto-correlation and intricate structure. Transforming $Y^*_t$ to $Z^*_t$ removes fine-scale periodic behavior, but $Z^*_t$ can still exhibit auto-correlation, especially in $Z_{t,s^*}^{*(1)}$. Therefore, we replace the fine-scale HMM within the hierarchical HMM with a CarHMM as shown in (fig \ref{fig:CarHHMM}).
%
\begin{figure}[ht]
	\centering
	\includegraphics[width=5in]{../Plots/CarHHMM.png}
	\caption{Graphical representation of a CarHHMM. The additional arrows representing auto-correlation between observations are shown in red for emphasis.}
	\label{fig:CarHHMM}
\end{figure}
%
Naturally, we refer to this model as the \textbf{CarHHMM}, and it can be implemented both with and without the short-time Fourier-transform (STFT). The likelihood of this model is still easy to calculate using the forward algorithm:
%
$$\calL_{\text{CarHHMM}}(y,z^*;\Theta,\Theta^*,\Gamma,\Gamma^*) = \delta P(y_1,z_1^*;\Theta,\Theta^*,\Gamma^*) \prod_{t=2}^T \Gamma P(y_t,z_t^*;\Theta,\Theta^*,\Gamma^*) \mathbf{1}_N$$
%
where:
%
\begin{align*}
P(y_t,z_t^*;\Theta,\Theta^*,\Gamma^*)  = \text{diag}\Big[&f^{(1)}(y_t)\calL_{\text{CarHMM}}\left(z_t^*;\Theta^{*(1)},\Gamma^{*(1)}\right), \ldots , \\
&f^{(N)}(y_t)\calL_{\text{CarHMM}}\left(z_t^*;\Theta^{*(N)},\Gamma^{*(N)}\right) \Big]
\end{align*}
%

### 2020-07-07 Report/Sections/simulation.tex
Updates from Overleaf

\section{Simulation Study}

@@

\begin{figure}[!ht]
	\centering
	\includegraphics[width=5in]{../Plots/sim_data.png}
	\caption{Simulated acceleration data for one dive. The color of the line corresponds to the true fine-scale state of the sub-dive process, while the color of the background corresponds to the true dive type of the simulated whale, where blue corresponds to dive type 1 and orange corresponds to dive type 2.}
	\label{fig:sim_data}
\end{figure}

@@

\begin{figure}[t!]
    \centering
    \begin{subfigure}[t]{1.0\textwidth}
        \centering
        \includegraphics[width=5in]{Plots/Posterior_Coarse_States.png}
        \caption{Coarse-scale hidden process}
    \end{subfigure}
    \newline
    \begin{subfigure}[t]{1.0\textwidth}
        \centering
        \includegraphics[width=5in]{Plots/Posterior_Fine_States.png}
        \caption{Fine-scale hidden process}
    \end{subfigure}
	\caption{Decoded state probabilities of each model. The color of the line corresponds to the true fine-scale state of the sub-dive process, while the color of the background corresponds to the true dive type of the simulated whale.}
	\label{fig:acc}
\end{figure}

%%% dive duration %%%
For the emission distributions of dive duration, estimates of standard error using the observed fisher information tended to be underestimates due to correlation between $\hat \mu$ and $\hat \sigma$. In addition, $\hat \sigma$ tends to underestimate $\sigma$ for all models, which is a finding consistent with properties of MLEs, especially because the sample size for a particular dive type is approximately 50 for each simulation. The CarHMM model in particular severely underestimates $\sigma$. This is likely due to the fact that dive duration follows a mixture of gammas rather than a gamma distribution, and the CarHMM does not have the machinary to capture this fact. A full table of parameter estimates for all models is shown in (table \ref{table:dive_duration}).

@@

\begin{figure}[!ht]
	\centering
	\includegraphics[width=5in]{../Plots/hmm_FV_MLE_density_dive_duration_-1.png}
	\caption{\textcolor{red}{KDE plot of $\hat \mu$ and $\hat \sigma$ for the dive duration emission distribution for the CarHMM. This plot is for Marie and Nancy to look over.}}
	\label{fig:MLE_dist}
\end{figure}

@@

\begin{table}[]

@@

>>>>

%\section{Simulation Study}

@@

\begin{figure}[ht]
	\centering
	\includegraphics[width=5in]{../Plots/sim_data.png}
	\caption{Simulated acceleration data for one dive. The color of the line corresponds to the true fine-scale state of the sub-dive process, while the color of the background corresponds to the true dive type of the simulated whale, where blue corresponds to dive type 1 and orange corresponds to dive type 2.}
	\label{fig:sim_data}
\end{figure}

@@

\begin{figure}[ht]
    \centering
    \begin{subfigure}[t]{1.0\textwidth}
        \centering
        \includegraphics[width=5in]{Plots/Posterior_Coarse_States.png}
        \caption{Coarse-scale hidden process}
    \end{subfigure}
    \newline
    \begin{subfigure}[t]{1.0\textwidth}
        \centering
        \includegraphics[width=5in]{Plots/Posterior_Fine_States.png}
        \caption{Fine-scale hidden process}
    \end{subfigure}
	\caption{Decoded state probabilities of each model. The color of the line corresponds to the true fine-scale state of the sub-dive process, while the color of the background corresponds to the true dive type of the simulated whale.}
	\label{fig:acc}
\end{figure}

%%% dive duration %%%
For the emission distributions of dive duration, estimates of standard error using the observed fisher information tended to be underestimates due to correlation between $\hat \mu$ and $\hat \sigma$. In addition, $\hat \sigma$ tends to underestimate $\sigma$ for all models, which is a finding consistent with properties of MLEs, especially because the sample size for a particular dive type is approximately 50 for each simulation. The CarHMM model in particular severely underestimates $\sigma$. This is likely due to the fact that dive duration follows a mixture of gammas rather than a gamma distribution, and the CarHMM does not have the machinery to capture this fact. A full table of parameter estimates for all models is shown in (table \ref{table:dive_duration}).

@@

\begin{figure}[ht]
	\centering
	\includegraphics[width=5in]{../Plots/hmm_FV_MLE_density_dive_duration_-1.png}
	\caption{\textcolor{red}{KDE plot of $\hat \mu$ and $\hat \sigma$ for the dive duration emission distribution for the CarHMM. This plot is for Marie and Nancy to look over.}}
	\label{fig:MLE_dist}
\end{figure}

@@

\begin{table}

@@

\scalebox{0.7}{
